Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-17 paper/agree.tex

This section introduces the AGREE contract specification language through the cyber-hardened system of \figref{fig:hardened}.
It details the specification of that top-level system interface.
It then discusses the specifications for the filter and the monitor added by the transforms.

This section then formalizes AGREE's assume-guarantee reasoning by defining the set of verification conditions that AGREE must check.
The verification conditions succinctly summarize what in means when AGREE proves that a system implementation satisfies it specification.
These conditions are useful in understanding how the semantics in synthesis differ from those in the assume-guarantee reasoning in \secref{sec:semantics}.

\subsection{Contract Specification}
\newsavebox{\sw}
\begin{lrbox}{\sw}
\begin{lstlisting}[style=agree,numbers=left]
eq req : bool = event(AutomationRequest);
eq avl : bool = event(AirVehicleLocation);
eq wp : bool = event(Waypoint);
eq strt: bool = event(Start);
eq alrt : bool = event(Alert);

assume "Automation requests are well-formed" :
  req => WELL_FORMED_AUTOMATION_REQUEST(AutomationRequest);
assume "Air vehicle locations are well-formed" :
  avl => WELL_FORMED_WAYPOINT(AirVehicleLocation);    
assume "One automation request in flight at a time" :
  true -> 
  (req => pre(Historically(not req) or Since(not req, strt)));

guarantee "Waypoints coincide with air vehicle locations":
  wp => avl;
guarantee "Starts include a new waypoint" :
  strt => wp;
guarantee "Waypoints are well-formed" : 
  wp => WELL_FORMED_WAYPOINT(Waypoint);
guarantee "Starts within one cycle of requests if not alerting" :
  (strt => ((not alrt) and req)) -> 
  (strt => ((not alrt) and (req or pre(req))));
guarantee "Alert if not started within one cycle of requests" :
    true -> ((pre(req and not strt) and not strt) => alrt);
guarantee "Once alerted always alerted" :
  not alrt or (Once(alrt) and alrt);
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \scalebox{0.62}{\usebox{\sw}}
  \end{center}
  \caption{The SW component contract.}
  \label{fig:sw}
\end{figure}

The AGREE specification for the SW component in the example of Section~\ref{sec:example} with the added cyber requirements is given in \figref{fig:sw}.
The AGREE specification language is a first-order predicate calculus that uses stream concepts, and operators, from the Lustre language \cite{10.1145/41625.41641}.
As with Lustre, the semantics are synchronous dataflow where the inputs, outputs, and expressions are data streams that comply with the input assumptions.
Contracts are evaluated in dependency order with inputs being propagated to outputs through all contracts until they stabilize; as such, the contracts, and thereby the top-level model, must be acyclic.\footnote{An apparent syntactic cycle, where a component is linked back to itself, may be broken temporally by inserting delay elements.}
Once the contracts have stabilized, the model takes a synchronous step to the next input data in the stream.
The semantics do not model computation or communication delay.
The output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

The AGREE model checker attempts to prove several properties of the top-level model being verified.
The first is that the output guarantees of each component implementing the system are strong enough to validate the input assumptions of any downstream component as well as to satisfy the guarantees of the output of the top-level component being verified (\ie, the system composition meets input assumptions at each input as well as the guarantees on the system output).
These properties are reported in the expanded lists in \figref{fig:example}(b) and \figref{fig:hardened}(b).
The next set of properties prove that the contract specifications for each component are \emph{self-consistent} \ie, a contract does not contradict itself.
These are the unexpanded results at the bottom of the figures.

Returning back to the contract of \figref{fig:sw}, it uses \texttt{eq} statements to define variables local to the contract specification.
For example, the \texttt{req} variable is equivalent to the \texttt{event(AutomationRequest)} expression.
In the AGREE semantics, there is an implicit \emph{event} input (or output) associated with every named event port in a component.
The semantics used here do not buffer these events so the implicit input (or output) is only a boolean value.
An \texttt{event} expression refers to that implicit input (or output) and is true when data is placed on the named port.
The system contract here states assumptions on well-formed input, followed by guarantees on properties about the output.

REPLACE ALL: 
The \emph{Alert if start is not bounded relative to a request} guarantee is an invariant on the expression \texttt{policy or since}, meaning that either the policy holds or the alert is sounding.
The \texttt{policy} is defined by two local values: \texttt{current} and \texttt{previous}.
The \texttt{current} value is asserted when in the current time step there is a request with a response, or there is no request and no response.

The value of \texttt{previous} in the current time step relies on values from the previous time step.
The \texttt{->} operator designates initialization, as the previous time step is undefined in the first step of the system.
The left operand to the operator is the initial value of \texttt{previous} at start, which in this example is \texttt{(req and not rsp)}, because seeing a request with no response is inconclusive in the first step of the system.
The right operand is the value of \texttt{previous} after the initial step.
Here the \texttt{pre} operator refers to the value of the expression \texttt{(req and not rsp)} in the prior time step, \texttt{previous} is true if the previous time step made a request without a matching response and the current time step has the matching response to that request with no new request.

The value of \texttt{since} in \emph{Alert if start is not bounded relative to a request} relies on its own value in the previous time step.
The intuitive reading of the expression is that the alert has been true since the time when it first sounded.
The first \texttt{alrt} sets \texttt{since} to true, and once the value of \texttt{since} is true, that value persists as long as \texttt{alrt} holds.
The \emph{Alert if start is not bounded relative to a request} guarantee defines one requirement of a cyber-hardened system implementation.
Together with the other guarantees, the contract models the expected input and output of the system as a whole.

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \includegraphics[scale=0.3]{dialogue.png}
    \end{tabular}
  \end{center}
  \caption{Wizard for automatically transforming the model with a filter.}
  \label{fig:dialogue}
\end{figure}

\newsavebox{\flt}
\begin{lrbox}{\flt}
\begin{lstlisting}[style=agree,numbers=left]
eq policy : bool = 
  WELL_FORMED_AUTOMATION_RESPONSE(Input);

guarantee Filter_Output
  "Filter output is well-formed" :
  if event(Input) and policy then 
    event(Output) and Output = Input
  else not event(Output);
\end{lstlisting}
\end{lrbox}

\newsavebox{\mntr}
\begin{lrbox}{\mntr}
\begin{lstlisting}[style=agree,numbers=left]
const is_latched : bool = true;
const MAX_LATENCY : int = 1;

eq rsp : bool = event(Response);
eq req : bool = event(Request);

eq isPending : bool = Since(not rsp, req and not rsp);
eq latency : int = 0 -> (if req then 0 else pre(latency) + 1);

eq policy : bool = (rsp => req) ->
                   (    (isPending => latency < MAX_LATENCY)   
                    and (rsp => (req or pre(isPending))));
eq alert : bool = (not policy) -> 
                  ((is_latched and pre(alert)) or not policy);

assume "One outstanding request at a time" :
  (true -> (req => not pre(isPending)));

guarantee "Alert port tracks alert variable" :
  event(Alert) = alert;
guarantee "Output if not alerted" :
  if (not(alert) and rsp) then
    event(Output) and (Output = Response)
  else
    not (event(Output));    
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.62}{\usebox{\flt}} \\
    \end{tabular}
  \end{center}
  \caption{Contract specification for high-assurance filter.}
  \label{fig:filter}
\end{figure}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.62}{\usebox{\mntr}} \\
    \end{tabular}
  \end{center}
  \caption{Contract specification for high-assurance monitor.}
  \label{fig:monitor}
\end{figure}

As noted previously, the original system fails to guarantee the cyber requirements.
BriefCASE provides two transformations to address the failing requirements: inserting a filter and inserting a monitor.
The component is added by selecting the connection in the model where the high-assurance component is to be added, and then choosing the appropriate transformation.
The system designer can provide transform configuration parameters in a wizard, as shown in \figref{fig:dialogue}.
The policy of the high-assurance component can be stated directly in the wizard, or it can be left blank.
In this example, the policy is specified as \texttt{WELL\_FORMED\_AUTOMATION\_RESPONSE(Input)}.
Additionally, because a transformation is ultimately driven by a cyber requirement, BriefCASE updates an embedded Resolute assurance case~\cite{resolute-destion}.
REMOVE: Resolute keeps track of the evidential artifacts necessary for supporting the requirement, and can be run at any time to determine whether those artifacts are valid.

The AGREE contract specification generated by the transform is shown in \figref{fig:filter}.
The guarantee is stylized for synthesis and completely defines the meaning of the output under every possible input.
The resulting AGREE specification for the monitor in this example is shown in \figref{fig:monitor}.
The \texttt{is\_latched} value makes the alert persistent, meaning that once the alert is raised, it is always raised.
This behavior is one of the several options available in the dialogue.
The definition for \texttt{policy} is taken by the system developer from the contract in \figref{fig:sw}.
As before, the guarantees for the outputs are autogenerated by the tool and completely define each output under every possible input.

@@

>>>>

This section formalizes AGREE's assume-guarantee reasoning by defining the set of verification conditions that AGREE must check.
The verification conditions succinctly summarize what in means when AGREE proves that a system implementation satisfies it specification.
These conditions are useful in understanding how the semantics in synthesis differ from those in the assume-guarantee reasoning in \secref{sec:semantics}.

This section then introduces the AGREE contract specification language through the cyber-hardened system of \figref{fig:hardened}.
It details the specification of that top-level system interface.
It then discusses the specifications for the filter and the monitor.
These two specifications are those that are synthesized to equivalent CakeML by SPLAT.

@@

\subsection{System Contract}
\input{sw-contract}

\subsection{Cyber-hardening the System Implementation}
\input{high-assurance-contracts}

### 2022-03-17 paper/sw-contract.tex

\newsavebox{\sw}
\begin{lrbox}{\sw}
\begin{lstlisting}[style=agree,numbers=left,escapeinside={*}{*}]
eq req : bool = event(AutomationRequest);*\label{line:sw-event-def-start}*
eq avl : bool = event(AirVehicleLocation);
eq wp : bool = event(Waypoint);
eq strt: bool = event(Start);
eq alrt : bool = event(Alert);*\label{line:sw-event-def-end}*

@@

All the named ports in the corresponding AADL component are in the scope of the specification, and there are additional implicit boolean \emph{event} inputs (or outputs) associated with event ports.
An \texttt{event} expression refers to that implicit input (or output) boolean value and is true when data present the named port and false otherwise.
\linesref{line:sw-event-def-start}{line:sw-event-def-end} create local variables that are true when data is present on the corresponding event ports for the component.
The local variables here are purely for convenience in writing the specification.

@@

The assumption on \lineref{line:sw-assume-3} uses a \emph{followed-by} operator, \texttt{->}, to \emph{guard} the \texttt{pre} operator at time 0.
A followed-by expression defines what happens at the first instance of the stream, time 0, and then what follows after. 
Here the assumption is \texttt{true} at time 0, and then takes on the truth value of the implication in all future instances.
The followed-by guards the \texttt{pre} because expressions are undefined before time 0 and the followed-by prevents the \texttt{pre} from being evaluated before time 1.
So after the first instance, the assumption depends on the presence of a request and the pre-state of the component.

The stream semantics in AGREE mean that it is possible to use PLTL operators that look back in time.
The \texttt{Historically} operator is the same as $\historically$ defined previously; it is true if the enclosed expression has been true in all moments up to the present one. 
The \texttt{Since} operator is true if the first expression has been true in all moments up to the present one since the second expression most recently became true.
The assumption on \lineref{line:sw-assume-3} is that if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it is after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}.

The guarantees on \linesref{line:sw-guarantee-1}{line:sw-guarantee-3} coincide events and assert well-formed output.
The guarantees on \linesref{line:sw-guarantee-4}{line:sw-guarantee-6} define temporal properties of the component.
\lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request.
The guarantee uses the assumption on \lineref{line:sw-assume-3} and does not check for two requests in a row without a start as the assumption precludes that input behavior.
The guarantee also does not force the start to take place, it only says that if it takes place, it is under the defined conditions.

@@

The \texttt{Once} operator is true if the enclosed expression has ever been true before the current moment.
\lineref{line:sw-guarantee-6} ensures that once alerted always alerted because either there has never been an alert, \texttt{\textbf{not} alrt}, or there has been an alert sometime in the past and so it is alerting in now, \texttt{\textbf{Once}(alrt) \textbf{and} alrt}.
These six guarantees define the behavior of the SW component under the three assumptions.

>>>>

\newsavebox{\sw}
\begin{lrbox}{\sw}
\begin{lstlisting}[style=agree,numbers=left]
eq req : bool = event(AutomationRequest);*\label{line:sw-event-def-start}*
eq avl : bool = event(AirVehicleLocation);
eq wp : bool = event(Waypoint);
eq strt: bool = event(Start);
eq alrt : bool = event(Alert);*\label{line:sw-event-def-end}*

@@

All the named ports in the corresponding AADL component are in the scope of the specification, and there are additional implicit boolean \emph{event} inputs (or outputs) associated with event ports.
An \texttt{event} expression refers to that implicit input (or output) boolean value and is true when data is present on the named port and false otherwise.
\linesref{line:sw-event-def-start}{line:sw-event-def-end} create local variables that are true when data is present on the corresponding event ports for the component.
The local variables here are purely for convenience in writing the specification.

@@

The assumption on \lineref{line:sw-assume-3} uses a \emph{followed-by} operator, \texttt{->}, to \emph{guard} the \texttt{pre} operator at time 0.
A followed-by expression defines what happens at the first instance of the stream, time 0, and then what follows after. 
Here the assumption is \texttt{true} at time 0 and then it is the truth value of the implication in all future instances.
The followed-by guards the \texttt{pre} because expressions are undefined before time 0 and the followed-by prevents the \texttt{pre} from being evaluated before time 1.
So after time 0, the assumption depends on the presence of a request and the pre-state of the component.

The stream semantics in AGREE mean that it is possible to use PLTL operators that look back in time.
The \texttt{Historically} operator is the same as $\historically$ defined previously; it is true if the enclosed expression has been true in all moments up to the present one. 
The \texttt{Since} operator is true if the first expression has been true in all moments up to the present one since the second expression most recently became true.
The assumption on \lineref{line:sw-assume-3} is that if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it is after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}--\emph{not request since start}.

The guarantees on \linesref{line:sw-guarantee-1}{line:sw-guarantee-3} coincide events and assert well-formed output.
The guarantees on \linesref{line:sw-guarantee-4}{line:sw-guarantee-6} define temporal properties of the component.
\lineref{line:sw-guarantee-4} insists that a start happens with a request or one step after a request.
The guarantee uses the assumption on \lineref{line:sw-assume-3} and does not check for two requests in a row without a start as the assumption precludes that input behavior.
It differentiates with the followed-by operator what is required at time 0, the start must coincide with the request, with what is required after time 0, the start must coincides with the request or is a response to a request one step earlier.
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.

@@

The \texttt{Once} operator is true if the enclosed expression has ever been true before the current moment.
\lineref{line:sw-guarantee-6} ensures that once alerted always alerted because either there has never been an alert, \texttt{\textbf{not} alrt}, or there has been an alert sometime in the past and so it is alerting now, \texttt{\textbf{Once}(alrt) \textbf{and} alrt}.
These six guarantees define the behavior of the SW component under the three assumptions.
